---
title: " Question 2b)"
author: "Candidate No. A13582 - Akithmi Wasalathanthri "
date: "2024-03-31"
output: html_document
---
"
****************************
*libraries to be installed*
****************************

```{r}
install.packages("ggplot2")
library(ggplot2)
library(gridExtra)
library(stats)
```
```{r}
install.packages("tidyverse")
library(tidyverse)
```

************************************************
#Data Cleaning and Preprocessing for part - 2b)
************************************************

*Reading the required datasets:*
(i)*The 2006 and 2007 merged dataset*
```{r}
merged_dataset <- read.csv("C:/Users/user/Desktop/progdata/mergedata2006n2007.csv")

```
(ii)*The plane dataset*
```{r}
dataset_planedata_ <- read.csv("C:/Users/user/Downloads/Programming/plane-data.csv")
dataset_planedata_
```

```{r}
colnames(dataset_planedata_)[colnames(dataset_planedata_) == "tailnum"] <- "TailNum"
colnames(dataset_planedata_)
```

*Creating a dataset combining the merged dataset and planedata dataset*
```{r}
# Merge the datasets
merged_planedata_ <- merge(merged_dataset, dataset_planedata_, by = "TailNum", all.x = TRUE)

names(merged_planedata_)[names(merged_planedata_) == "year"] <- "YearOfManufacture"

str(merged_planedata_)

q2b_dataset <- merged_planedata_[c("TailNum", "Year", "YearOfManufacture", "ArrDelay", "DepDelay", "CarrierDelay", "WeatherDelay", "NASDelay", "SecurityDelay", "LateAircraftDelay")]
head(q2b_dataset)
```


```{r}
# Check the data type of the "YearOfManufacture" column
class(q2b_dataset$YearOfManufacture)

# Check for missing values
any(is.na(q2b_dataset$YearOfManufacture))


# Count the number of missing values in the entire dataset
num_missing <- sum(is.na(q2b_dataset))

# Print the number of missing values
print(paste("The number of missing values in the dataset is:", num_missing))

# Remove rows with missing values
q2b_dataset_clean <- na.omit(q2b_dataset)

# Count the number of missing values in the cleaned dataset
num_missing_clean <- sum(is.na(q2b_dataset_clean))

# Print the number of missing values
print(paste("The number of missing values in the cleaned dataset is:", num_missing_clean))
```

```{r}
q2b_arr_dataset <- q2b_dataset_clean[, c("TailNum", "Year", "YearOfManufacture", "ArrDelay")]

# Display the first few rows
head(q2b_arr_dataset)
q2b_dep_dataset <- q2b_dataset_clean[, c("TailNum", "Year", "YearOfManufacture", "DepDelay")]

# Display the first few rows
head(q2b_dep_dataset)

```


```{r}
# Check for null values in q2b_arr_dataset
null_arr <- colSums(is.na(q2b_arr_dataset))
print("Null values in q2b_arr_dataset:")
print(null_arr)

# Check for null values in q2b_dep_dataset
null_dep <- colSums(is.na(q2b_dep_dataset))
print("Null values in q2b_dep_dataset:")
print(null_dep)

```

```{r}
# Removing early arrivals
q2b_arr_dataset <- q2b_arr_dataset[q2b_arr_dataset$ArrDelay >= 0, ]

# Removing early departures
q2b_dep_dataset <- q2b_dep_dataset[q2b_dep_dataset$DepDelay >= 0, ]

# Display the first few rows of cleaned datasets
head(q2b_arr_dataset)
head(q2b_dep_dataset)
```

```{r}
# Convert Year and YearOfManufacture columns to numeric
q2b_arr_dataset$Year <- as.numeric(q2b_arr_dataset$Year)
q2b_arr_dataset$YearOfManufacture <- as.numeric(q2b_arr_dataset$YearOfManufacture)

# Calculate plane age for q2b_arr_dataset
q2b_arr_dataset$PlaneAge <- q2b_arr_dataset$Year - q2b_arr_dataset$YearOfManufacture

# Convert Year and YearOfManufacture columns to numeric
q2b_dep_dataset$Year <- as.numeric(q2b_dep_dataset$Year)
q2b_dep_dataset$YearOfManufacture <- as.numeric(q2b_dep_dataset$YearOfManufacture)

# Calculate plane age for q2b_arr_dataset
q2b_dep_dataset$PlaneAge <- q2b_dep_dataset$Year - q2b_dep_dataset$YearOfManufacture


```


```{r}
# Print the number of rows in q2b_arr_dataset
print(paste("Number of rows in q2b_arr_dataset:", nrow(q2b_arr_dataset)))
# Print the number of rows in q2b_dep_dataset
print(paste("Number of rows in q2b_dep_dataset:", nrow(q2b_dep_dataset)))

# Count the number of rows with YearOfManufacture equal to 0 in q2b_arr_dataset
num_rows_year_0_arr <- sum(q2b_arr_dataset$YearOfManufacture == 0)
print(paste("Number of rows with YearOfManufacture equal to 0 in q2b_arr_dataset:", num_rows_year_0_arr))

# Count the number of rows with YearOfManufacture equal to 0 in q2b_dep_dataset
num_rows_year_0_dep <- sum(q2b_dep_dataset$YearOfManufacture == 0)
print(paste("Number of rows with YearOfManufacture equal to 0 in q2b_dep_dataset:", num_rows_year_0_dep))

# Remove rows with YearOfManufacture equal to 0 from q2b_arr_dataset
q2b_arr_dataset <- q2b_arr_dataset[q2b_arr_dataset$YearOfManufacture != 0, ]
print(paste("Number of rows in q2b_arr_dataset:", nrow(q2b_arr_dataset)))
# Remove rows with YearOfManufacture equal to 0 from q2b_dep_dataset
q2b_dep_dataset <- q2b_dep_dataset[q2b_dep_dataset$YearOfManufacture != 0, ]
print(paste("Number of rows in q2b_dep_dataset:", nrow(q2b_dep_dataset)))

```

```{r}
# Check for plane age values of -1 in q2b_arr_dataset
q2b_arr_dataset$PlaneAge[q2b_arr_dataset$PlaneAge == -1] <- NA
# Count the number of NULL values in the PlaneAge column for q2b_arr_dataset
num_null_arr <- sum(is.na(q2b_arr_dataset$PlaneAge))
print(paste("Number of NULL values in q2b_arr_dataset:", num_null_arr))

# Check for plane age values of -1 in q2b_dep_dataset
q2b_dep_dataset$PlaneAge[q2b_dep_dataset$PlaneAge == -1] <- NA
# Count the number of NULL values in the PlaneAge column for q2b_dep_dataset
num_null_dep <- sum(is.na(q2b_dep_dataset$PlaneAge))
print(paste("Number of NULL values in q2b_dep_dataset:", num_null_dep))

# Remove rows with NULL values in the PlaneAge column for q2b_arr_dataset
q2b_arr_dataset <- na.omit(q2b_arr_dataset)

# Remove rows with NULL values in the PlaneAge column for q2b_dep_dataset
q2b_dep_dataset <- na.omit(q2b_dep_dataset)

```

```{r}
# Print the number of rows in q2b_arr_dataset
print(paste("Number of rows in q2b_arr_dataset:", nrow(q2b_arr_dataset)))
# Print the number of rows in q2b_dep_dataset
print(paste("Number of rows in q2b_dep_dataset:", nrow(q2b_dep_dataset)))
```

**********************************************************************

***********
#Part - A
***********

*Reading the two cleaned datasets*
```{r}
q2b_arr_dataset_cleaned <- read.csv("C:/Users/user/Desktop/progdata/q2b_arr_dataset_cleaned.csv")
head(q2b_arr_dataset_cleaned)
q2b_dep_dataset_cleaned <- read.csv("C:/Users/user/Desktop/progdata/q2b_dep_dataset_cleaned.csv")
head(q2b_dep_dataset_cleaned)
```

#calculating the Mean Arrival Delay*

*For 2006 dataset*
```{r}
q2b_arr_dataset_2006 <- subset(q2b_arr_dataset_cleaned, Year == 2006)
yearly_arrival_data_2006 <- aggregate(ArrDelay ~ PlaneAge, data = q2b_arr_dataset_2006, FUN = mean)
cat("Mean arrival delay for each plane age in 2006:\n")
print(head(yearly_arrival_data_2006))
```

*For 2007 dataset*
```{r}
q2b_arr_dataset_2007 <- subset(q2b_arr_dataset_cleaned, Year == 2007)
yearly_arrival_data_2007 <- aggregate(ArrDelay ~ PlaneAge, data = q2b_arr_dataset_2007, FUN = mean)
cat("\nMean arrival delay for each plane age in 2007:\n")
print(head(yearly_arrival_data_2007))
```


```{r}
# Plot for 2006
plot_2006 <- ggplot(data = yearly_arrival_data_2006, aes(x = PlaneAge, y = ArrDelay)) +
  geom_point(color = "red", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Relationship between Plane Age and Mean Arrival Delay (2006)",
       x = "Plane Age", y = "Mean Arrival Delay") +
  theme_minimal()

# Plot for 2007
plot_2007 <- ggplot(data = yearly_arrival_data_2007, aes(x = PlaneAge, y = ArrDelay)) +
  geom_point(color = "green", alpha = 0.7) +
  geom_smooth(method = "lm", color = "green", se = FALSE) +
  labs(title = "Relationship between Plane Age and Mean Arrival Delay (2007)",
       x = "Plane Age", y = "Mean Arrival Delay") +
  theme_minimal()

grid.arrange(plot_2006, plot_2007, ncol = 1)
```

```{r}
# Calculate Pearson correlation coefficient for 2006
corr_test_2006 <- cor.test(yearly_arrival_data_2006$PlaneAge, yearly_arrival_data_2006$ArrDelay)

# Print the correlation coefficient  for 2006
cat("Correlation between Plane Age and Mean Arrival Delay (2006): ", corr_test_2006$estimate, "\n")

# Calculate Pearson correlation coefficient for 2007
corr_test_2007 <- cor.test(yearly_arrival_data_2007$PlaneAge, yearly_arrival_data_2007$ArrDelay)

# Print the correlation coefficient for 2007
cat("\nCorrelation between Plane Age and Mean Arrival Delay (2007): ", corr_test_2007$estimate, "\n")
```

#Calculating Mean Departure Delay
*For 2006 dataset*
```{r}
# Filter dataset for 2006
q2b_dep_dataset_2006 <- subset(q2b_dep_dataset_cleaned, Year == 2006)

# Calculate mean departure delay for each plane age in 2006
yearly_departure_data_2006 <- aggregate(DepDelay ~ PlaneAge, data = q2b_dep_dataset_2006, FUN = mean)
print(head(yearly_departure_data_2006))
```
*For 2007 dataset*
```{r}
# Filter dataset for 2007
q2b_dep_dataset_2007 <- subset(q2b_dep_dataset_cleaned, Year == 2007)

# Calculate mean departure delay for each plane age in 2007
yearly_departure_data_2007 <- aggregate(DepDelay ~ PlaneAge, data = q2b_dep_dataset_2007, FUN = mean)
print(head(yearly_departure_data_2007))
```

```{r}
# Create subplots
plot_2006 <- ggplot(data = yearly_departure_data_2006, aes(x = PlaneAge, y = DepDelay)) +
  geom_point(color = "red", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Relationship between Plane Age and Mean Departure Delay (2006)",
       x = "Plane Age", y = "Mean Departure Delay") +
  theme_minimal()

plot_2007 <- ggplot(data = yearly_departure_data_2007, aes(x = PlaneAge, y = DepDelay)) +
  geom_point(color = "green", alpha = 0.7) +
  geom_smooth(method = "lm", color = "green", se = FALSE) +
  labs(title = "Relationship between Plane Age and Mean Departure Delay (2007)",
       x = "Plane Age", y = "Mean Departure Delay") +
  theme_minimal()

grid.arrange(plot_2006, plot_2007, ncol = 1)

```

```{r}
# Calculate Pearson correlation coefficient and p-value for 2006
corr_test_2006 <- cor.test(yearly_departure_data_2006$PlaneAge, yearly_departure_data_2006$DepDelay)

# Print the correlation coefficient for 2006
cat("Correlation between Plane Age and Mean Departure Delay (2006): ", corr_test_2006$estimate, "\n")


# Calculate Pearson correlation coefficient and p-value for 2007
corr_test_2007 <- cor.test(yearly_departure_data_2007$PlaneAge, yearly_departure_data_2007$DepDelay)

# Print the correlation coefficient and p-value for 2007
cat("\nCorrelation between Plane Age and Mean Departure Delay (2007): ", corr_test_2007$estimate, "\n")

```

```{r}

# Plot for 2006
plot_2006 <- ggplot() +
  geom_line(data = yearly_arrival_data_2006, aes(x = PlaneAge, y = ArrDelay, color = "Mean Arrival Delay (2006)")) +
  geom_line(data = yearly_departure_data_2006, aes(x = PlaneAge, y = DepDelay, color = "Mean Departure Delay (2006)")) +
  labs(title = "Mean Arrival and Departure Delay by Plane Age (2006)",
       x = "Plane Age", y = "Mean Delay (minutes)") +
  scale_color_manual(values = c("red", "darkgreen")) +
  theme_minimal()

# Plot for 2007
plot_2007 <- ggplot() +
  geom_line(data = yearly_arrival_data_2007, aes(x = PlaneAge, y = ArrDelay, color = "Mean Arrival Delay (2007)")) +
  geom_line(data = yearly_departure_data_2007, aes(x = PlaneAge, y = DepDelay, color = "Mean Departure Delay (2007)")) +
  labs(title = "Mean Arrival and Departure Delay by Plane Age (2007)",
       x = "Plane Age", y = "Mean Delay (minutes)") +
  scale_color_manual(values = c("red", "darkgreen")) +
  theme_minimal()

# Arrange plots in a grid layout
grid.arrange(plot_2006, plot_2007, ncol = 1)

```

```{r}
# Plot for 2006
plot_2006 <- ggplot() +
  geom_line(data = yearly_arrival_data_2006, aes(x = PlaneAge, y = ArrDelay, color = "Mean Arrival Delay (2006)"), size = 0.7) +
  geom_line(data = yearly_departure_data_2006, aes(x = PlaneAge, y = DepDelay, color = "Mean Departure Delay (2006)"), size = 0.7) +
  labs(title = "Mean Arrival and Departure Delay by Plane Age (2006)",
       x = "Plane Age", y = "Mean Delay (minutes)") +
  scale_color_manual(values = c("red", "darkgreen")) +
  theme_minimal()

# Plot for 2007
plot_2007 <- ggplot() +
  geom_line(data = yearly_arrival_data_2007, aes(x = PlaneAge, y = ArrDelay, color = "Mean Arrival Delay (2007)"), size = 0.7) +
  geom_line(data = yearly_departure_data_2007, aes(x = PlaneAge, y = DepDelay, color = "Mean Departure Delay (2007)"), size = 0.7) +
  labs(title = "Mean Arrival and Departure Delay by Plane Age (2007)",
       x = "Plane Age", y = "Mean Delay (minutes)") +
  scale_color_manual(values = c("red", "darkgreen")) +
  theme_minimal()

grid.arrange(plot_2006, plot_2007, ncol = 1)
```
*******************************************************************

***********
# Part - B 
***********

```{r}
q2b_dataset <- read.csv("C:/Users/user/Desktop/progdata/q2b_dataset.csv")
head(q2b_dataset)
na_count <- colSums(is.na(q2b_dataset))
# Displaying the number of null values for each column
print(na_count)
```

```{r}
#Sort the q2dataset by YearOfManufacture in ascending order
q2b_dataset <- q2b_dataset[order(q2b_dataset$YearOfManufacture), ]
# Display the sorted dataframe
print(q2b_dataset)
```

```{r}
# Drop columns 'ArrDelay' and 'DepDelay'
q2b_delaytypes_only <- q2b_dataset[, !(names(q2b_dataset) %in% c('ArrDelay', 'DepDelay'))]

# Drop rows with null values and YearOfManufacture equal to zero
q2b_delaytypes_only <- q2b_delaytypes_only[complete.cases(q2b_delaytypes_only) & q2b_delaytypes_only$YearOfManufacture != 0, ]

# Print the resulting dataframe
print(q2b_delaytypes_only)
```

```{r}
# Calculate plane age
q2b_delaytypes_only$PlaneAge <- q2b_delaytypes_only$Year - q2b_delaytypes_only$YearOfManufacture

# Drop rows with PlaneAge less than 0
q2b_delaytypes_only <- q2b_delaytypes_only[q2b_delaytypes_only$PlaneAge >= 0, ]

# Print the resulting dataframe
print(q2b_delaytypes_only)
```

```{r}
# Create a new column 'PlaneAgeCondition' with initial value 'old'
q2b_delaytypes_only$PlaneAgeCondition <- 'old'

# Update 'PlaneAgeCondition' to 'new' for rows where 'PlaneAge' is less than or equal to 20
q2b_delaytypes_only$PlaneAgeCondition[q2b_delaytypes_only$PlaneAge <= 20] <- 'new'

# Print the dataset
print(q2b_delaytypes_only)
```

#seperating the dataset according to the 5 different delay types*
```{r}
# Filter rows with CarrierDelay greater than or equal to 1
q2b_carrier_delay <- subset(q2b_delaytypes_only, CarrierDelay >= 1, select=c("CarrierDelay", "PlaneAge", "Year", "PlaneAgeCondition"))
print(q2b_carrier_delay)

# Filter rows with WeatherDelay greater than or equal to 1
q2b_weather_delay <- subset(q2b_delaytypes_only, WeatherDelay >= 1, select=c("WeatherDelay", "PlaneAge", "Year", "PlaneAgeCondition"))
print(q2b_weather_delay)

# Filter rows with NASDelay greater than or equal to 1
q2b_NAS_delay <- subset(q2b_delaytypes_only, NASDelay >= 1, select=c("NASDelay", "PlaneAge", "Year", "PlaneAgeCondition"))
print(q2b_NAS_delay)

# Filter rows with SecurityDelay greater than or equal to 1
q2b_security_delay <- subset(q2b_delaytypes_only, SecurityDelay >= 1, select=c("SecurityDelay", "PlaneAge", "Year", "PlaneAgeCondition"))
print(q2b_security_delay)

# Filter rows with LateAircraftDelay greater than or equal to 1
q2b_lateaircraft_delay <- subset(q2b_delaytypes_only, LateAircraftDelay >= 1, select=c("LateAircraftDelay", "PlaneAge", "Year", "PlaneAgeCondition"))
print(q2b_lateaircraft_delay)
```

*Seperating the delays according to the delaytype*
```{r}
# Subset data frames for each year
q2b_carrier_delay_2006 <- subset(q2b_carrier_delay, Year == 2006)
q2b_carrier_delay_2007 <- subset(q2b_carrier_delay, Year == 2007)

q2b_weather_delay_2006 <- subset(q2b_weather_delay, Year == 2006)
q2b_weather_delay_2007 <- subset(q2b_weather_delay, Year == 2007)

q2b_NAS_delay_2006 <- subset(q2b_NAS_delay, Year == 2006)
q2b_NAS_delay_2007 <- subset(q2b_NAS_delay, Year == 2007)

q2b_security_delay_2006 <- subset(q2b_security_delay, Year == 2006)
q2b_security_delay_2007 <- subset(q2b_security_delay, Year == 2007)

q2b_lateaircraft_delay_2006 <- subset(q2b_lateaircraft_delay, Year == 2006)
q2b_lateaircraft_delay_2007 <- subset(q2b_lateaircraft_delay, Year == 2007)
```

*Seperating the variables according to the delay type and year*
```{r}

carrier_delay_2006 <- q2b_carrier_delay_2006$CarrierDelay
carrier_delay_2007 <- q2b_carrier_delay_2007$CarrierDelay

weather_delay_2006 <- q2b_weather_delay_2006$WeatherDelay
weather_delay_2007 <- q2b_weather_delay_2007$WeatherDelay

NAS_delay_2006 <- q2b_NAS_delay_2006$NASDelay
NAS_delay_2007 <- q2b_NAS_delay_2007$NASDelay

security_delay_2006 <- q2b_security_delay_2006$SecurityDelay
security_delay_2007 <- q2b_security_delay_2007$SecurityDelay

lateaircraft_delay_2006 <- q2b_lateaircraft_delay_2006$LateAircraftDelay
lateaircraft_delay_2007 <- q2b_lateaircraft_delay_2007$LateAircraftDelay
```

```{r,fig.width= 10, fig.height= 10}

# Create a new plot with specified size
plot.new()
plot.window(xlim=c(0.5, 5.8), ylim=c(0, 140), xlab="", ylab="Delay Time (minutes)", main="Comparison of Delay Types between 2006 and 2007")

# Define positions for each boxplot
positions_2006 <- c(1, 2, 3, 4, 5)
positions_2007 <- c(1.3, 2.3, 3.3, 4.3, 5.3)

# Plot boxplots for 2006 data with different colors and smaller width, without outliers
boxplot(carrier_delay_2006, weather_delay_2006, NAS_delay_2006, security_delay_2006, lateaircraft_delay_2006,
        at=positions_2006, 
        add=TRUE, 
        col="darkgreen", 
        boxwex=0.1,
        outline=FALSE)

# Plot boxplots for 2007 data with different colors and smaller width, without outliers
boxplot(carrier_delay_2007, weather_delay_2007, NAS_delay_2007, security_delay_2007, lateaircraft_delay_2007,
        at=positions_2007, 
        add=TRUE, 
        col="red",
        boxwex=0.1,
        outline=FALSE)

# Add x-axis labels and ticks
# Add x-axis labels and ticks with adjusted font size
axis(1, at=c(positions_2006, positions_2007), labels=rep(c('Carrier Delay', 'Weather Delay', 'NAS Delay', 'Security Delay', 'Late Aircraft Delay'), 2), las=2, cex.axis=0.6)

# Add legend
legend("topright", legend=c("2006", "2007"), fill=c("darkgreen", "red"))

# Add title
title("Comparison of Delay Types between 2006 and 2007")
``` 

************
# Year 2006
************
*calculating the mean for each delay type with their plan age condition - 2006 dataset*
```{r}
carrier_delay_avg_2006 <- aggregate(q2b_carrier_delay_2006$CarrierDelay, 
                                    by = list(q2b_carrier_delay_2006$PlaneAgeCondition), FUN = mean)
names(carrier_delay_avg_2006) <- c("PlaneAgeCondition", "CarrierDelay")

weather_delay_avg_2006 <- aggregate(q2b_weather_delay_2006$WeatherDelay, 
                                    by = list(q2b_weather_delay_2006$PlaneAgeCondition), 
                                    FUN = mean)
names(weather_delay_avg_2006) <- c("PlaneAgeCondition", "WeatherDelay")

NAS_delay_avg_2006 <- aggregate(q2b_NAS_delay_2006$NASDelay, 
                                by = list(q2b_NAS_delay_2006$PlaneAgeCondition), 
                                FUN = mean)
names(NAS_delay_avg_2006) <- c("PlaneAgeCondition", "NASDelay")

security_delay_avg_2006 <- aggregate(q2b_security_delay_2006$SecurityDelay, 
                                     by = list(q2b_security_delay_2006$PlaneAgeCondition), 
                                     FUN = mean)
names(security_delay_avg_2006) <- c("PlaneAgeCondition", "SecurityDelay")

lateaircraft_delay_avg_2006 <- aggregate(q2b_lateaircraft_delay_2006$LateAircraftDelay, 
                                         by = list(q2b_lateaircraft_delay_2006$PlaneAgeCondition), 
                                         FUN = mean)
names(lateaircraft_delay_avg_2006) <- c("PlaneAgeCondition", "LateAircraftDelay")
```


*merging the mean for each delay types for seperate years*
```{r}
# Merge all delay types for 2006
merged_delay_types_2006 <- merge(carrier_delay_avg_2006, weather_delay_avg_2006, by = 'PlaneAgeCondition')
merged_delay_types_2006 <- merge(merged_delay_types_2006, NAS_delay_avg_2006, by = 'PlaneAgeCondition')
merged_delay_types_2006 <- merge(merged_delay_types_2006, security_delay_avg_2006, by = 'PlaneAgeCondition')
merged_delay_types_2006 <- merge(merged_delay_types_2006, lateaircraft_delay_avg_2006, by = 'PlaneAgeCondition')
merged_delay_types_2006
```

```{r, fig.width= 10}
# Sample data
merged_delay_types_2006 <- data.frame(
  PlaneAgeCondition = c('new', 'old'),
  CarrierDelay = c(35.36170, 32.90817),
  WeatherDelay = c(44.16256, 34.85578),
  NASDelay = c(28.36564, 22.19685),
  SecurityDelay = c(22.00550, 18.20117),
  LateAircraftDelay = c(45.91039, 36.37087)
)

# Extracting delay values for each PlaneAgeCondition
new_values <- as.matrix(merged_delay_types_2006[merged_delay_types_2006$PlaneAgeCondition == 'new', -1])
old_values <- as.matrix(merged_delay_types_2006[merged_delay_types_2006$PlaneAgeCondition == 'old', -1])

# Colors for bars
colors <- c('#800020', 'darkgreen')

# Plotting the bar chart
bar_width <- 0.35
index <- seq_along(names(new_values))
barplot(
  rbind(new_values, old_values),
  beside = TRUE,
  names.arg = names(new_values),
  col = colors,
  main = 'Delay Time for each delay type - 2006: New vs Old',
  xlab = 'Delay Types',
  ylab = 'Mean Delay Time in Minutes',
  legend.text = c('New', 'Old'),
  args.legend = list(x = 'topleft', bty = 'n')
)
```


***********
#Year 2007
***********
*calculating the mean for each delay type with their plan age condition - 2007 dataset*
```{r}
carrier_delay_avg_2007 <- aggregate(q2b_carrier_delay_2007$CarrierDelay, 
                                    by = list(q2b_carrier_delay_2007$PlaneAgeCondition), FUN = mean)
names(carrier_delay_avg_2007) <- c("PlaneAgeCondition", "CarrierDelay")

weather_delay_avg_2007 <- aggregate(q2b_weather_delay_2007$WeatherDelay, 
                                    by = list(q2b_weather_delay_2007$PlaneAgeCondition), 
                                    FUN = mean)
names(weather_delay_avg_2007) <- c("PlaneAgeCondition", "WeatherDelay")

NAS_delay_avg_2007 <- aggregate(q2b_NAS_delay_2007$NASDelay, 
                                by = list(q2b_NAS_delay_2007$PlaneAgeCondition), 
                                FUN = mean)
names(NAS_delay_avg_2007) <- c("PlaneAgeCondition", "NASDelay")

security_delay_avg_2007 <- aggregate(q2b_security_delay_2007$SecurityDelay, 
                                     by = list(q2b_security_delay_2007$PlaneAgeCondition), 
                                     FUN = mean)
names(security_delay_avg_2007) <- c("PlaneAgeCondition", "SecurityDelay")

lateaircraft_delay_avg_2007 <- aggregate(q2b_lateaircraft_delay_2007$LateAircraftDelay, 
                                         by = list(q2b_lateaircraft_delay_2007$PlaneAgeCondition), 
                                         FUN = mean)
names(lateaircraft_delay_avg_2007) <- c("PlaneAgeCondition", "LateAircraftDelay")

```

*mergin the above dataset with plane age condition - 2007*
```{r}
merged_delay_types_2007 <- merge(carrier_delay_avg_2007, weather_delay_avg_2007, by = 'PlaneAgeCondition')
merged_delay_types_2007 <- merge(merged_delay_types_2007, NAS_delay_avg_2007, by = 'PlaneAgeCondition')
merged_delay_types_2007 <- merge(merged_delay_types_2007, security_delay_avg_2007, by = 'PlaneAgeCondition')
merged_delay_types_2007 <- merge(merged_delay_types_2007, lateaircraft_delay_avg_2007, by = 'PlaneAgeCondition')
merged_delay_types_2007
```

```{r, fig.width= 10}
# Sample data
merged_delay_types_2007 <- data.frame(
  PlaneAgeCondition = c('new', 'old'),
  CarrierDelay = c(34.47172, 32.56365),
  WeatherDelay = c(44.10177, 36.29995),
  NASDelay = c(27.78155, 23.47370),
  SecurityDelay = c(19.25471, 15.48352),
  LateAircraftDelay = c(45.68296, 39.89857)
)

# Extracting delay values for each PlaneAgeCondition
new_values <- as.matrix(merged_delay_types_2007[merged_delay_types_2007$PlaneAgeCondition == 'new', -1])
old_values <- as.matrix(merged_delay_types_2007[merged_delay_types_2007$PlaneAgeCondition == 'old', -1])

# Colors for bars
colors <- c('#800020', 'darkgreen')

# Plotting the bar chart
bar_width <- 0.35
index <- seq_along(names(new_values))
barplot(
  rbind(new_values, old_values),
  beside = TRUE,
  names.arg = names(new_values),
  col = colors,
  main = 'Delay Time for each delay type - 2007: New vs Old',
  xlab = 'Delay Types',
  ylab = 'Mean Delay Time in Minutes',
  legend.text = c('New', 'Old'),
  args.legend = list(x = 'topleft', bty = 'n')
)
```

*****
END
*****























































































."

