---
title: "Question 2a)"
author: "Candidate No. A13582 - Akithmi Wasalathanthri"
date: "2024-03-24"
output: html_document
---
"
*********************
*"Question - 2a)"*
*********************

*"Importing the necessary libraries for question 2 part (a)"*
```{r}
library(readr) 
# "For reading CSV files"
library(dplyr) 
# "For data manipulation"
library(ggplot2) 
# "For data visualization"
```


```{r}
# "Read the merged data set for 2006 and 2007"
merged_dataset <- read.csv("C:/Users/user/Desktop/progdata/mergedata2006n2007.csv")
```

```{r}
# "Checking the data types in the merged data set"
str(merged_dataset)

# "Changing the data type of the 'CRSDepTime' column to integer"
merged_dataset$CRSDepTime <- as.integer(merged_dataset$CRSDepTime)

# "Rechecking data types of the merged data set"
str(merged_dataset)
```

******************************************************
*"Cleaning the data set seperately for Arrival Delay"*
******************************************************

```{r}
# "Extracting the desired columns for question 2 part (a) for arrival delay"
q2a_arr_dataset <- merged_dataset[, c("Year", "Month", "DayOfWeek", "CRSDepTime", "ArrDelay")]

# "Displaying the first few rows of the data set
head(q2a_arr_dataset)

# "Checking for any null values in the data set 'q2a_arr_dataset'"
colSums(is.na(q2a_arr_dataset))

# "Dropping the null values in q2a_arr_dataset"
q2a_arr_dataset <- na.omit(q2a_arr_dataset)

# "Displaying the first few rows after removing null values"
head(q2a_arr_dataset)

# "Re-checking for any null values that weren't removed"
colSums(is.na(q2a_arr_dataset))

# "Removing early arrivals from the data set"
q2a_arr_dataset <- q2a_arr_dataset[q2a_arr_dataset$ArrDelay >= 0, ]
head(q2a_arr_dataset)

```

************************************************************
*" Cleaning the data set separately for Departure Delay"*
************************************************************

```{r}
# "Extracting the desired columns for question 2 part (a) for departure delay"
q2a_dep_dataset <- merged_dataset[, c("Year", "Month", "DayOfWeek", "CRSDepTime", "DepDelay")]

# "Checking for any null values in the data set 'q2a_dep_dataset'"
colSums(is.na(q2a_dep_dataset))

# "Dropping the null values in q2a_dep_dataset"
q2a_dep_dataset <- na.omit(q2a_dep_dataset)

#"Re-checking for any null values that weren't removed"
colSums(is.na(q2a_dep_dataset))

# "Removing early departures from the data set"
q2a_dep_dataset <- q2a_dep_dataset[q2a_dep_dataset$DepDelay >= 0, ]
head(q2a_dep_dataset)
```

***********************************************************************************
*"(i) Analyzing the distribution of arrival delay and departure delay variables"*
***********************************************************************************

```{r}
# "Extracting only the Arrival Delay data"
delay_arr <- q2a_arr_dataset$ArrDelay

# "Extracting only the Departure Delay data"
delay_dep <- q2a_dep_dataset$DepDelay

# "Filtering the data needed by year only for the year 2006"
delayfor_arr_06 <- q2a_arr_dataset[q2a_arr_dataset$Year == 2006, ]$ArrDelay
delayfor_dep_06 <- q2a_dep_dataset[q2a_dep_dataset$Year == 2006, ]$DepDelay

```

*"Plotting a box plot and a violin plot to show the distributions with outliers included for year 2006"*

```{r}
# "Install and load the 'vioplot' package"
install.packages("vioplot")
library(vioplot)

# "Creating a boxplot and violin plot for 2006"
par(mfrow = c(1, 2), mar = c(5, 5, 2, 2))
boxplot(delayfor_arr_06, delayfor_dep_06, names = c("Arrival Delay", "Dep Delay"), 
        main = "Box Plot - 2006", 
        xlab = "Delay Type", ylab = "Delay Time (minutes)", col = c("navy", "orange"))
vioplot(delayfor_arr_06, delayfor_dep_06, names = c("Arrival Delay", "Dep Delay"), 
        main = "Violin Plot - 2006 ", 
        xlab = "Delay Type", ylab = "Density", col = c("navy", "orange"))

```

 *"Plotting a box plot and a violin plot to show the distributions with outliers included for year 2007"*
 
```{r}
# "Filtering the data needed by year only for the year 2007"
delayfor_arr_2007 <- q2a_arr_dataset[q2a_arr_dataset$Year == 2007, ]$ArrDelay
delayfor_dep_2007 <- q2a_dep_dataset[q2a_dep_dataset$Year == 2007, ]$DepDelay

# "Creating a boxplot and violin plot for 2006"
par(mfrow = c(1, 2), mar = c(5, 5, 2, 2))
boxplot(delayfor_arr_2007, delayfor_dep_2007, names = c("Arrival Delay", "Dep Delay"), 
        main = "Box Plot - 2007", 
        xlab = "Delay Type", ylab = "Delay Time (minutes)", col = c("navy", "orange"))
vioplot(delayfor_arr_06, delayfor_dep_2007, names = c("Arrival Delay", "Dep Delay"), 
        main = "Violin Plot - 2007", 
        xlab = "Delay Type", ylab = "Density", col = c("navy", "orange"))
```

 *"Comparing the distributions using box plots only by each year for 2006 and 2007"*

```{r}
par(mfrow = c(1, 2), mar = c(5, 5, 2, 2))
boxplot(delayfor_arr_06, delayfor_dep_06, names = c("Arrival Delay", "Dep Delay"), 
        main = "Box Plot - 2006", 
        xlab = "Delay Type", ylab = "Delay Time (minutes)", col = c("#08526A", "#08526A"), 
        outline = FALSE, boxwex = 0.5)

boxplot(delayfor_arr_2007, delayfor_dep_2007, names = c("Arrival Delay", "Dep Delay"), 
        main = "Box Plot - 2007", 
        xlab = "Delay Type", ylab = "Delay Time (minutes)", col = c("#e43c6f", "#e43c6f"), 
        outline = FALSE, boxwex = 0.5)


```

*****************************************************************************
*"(ii) Evaluating the Best time of the day in a week to travel (by hour)"*
*****************************************************************************

```{r}
# "Grouping the Scheduled Arrival Time into intervals or bins to analyze arrival delays data set separately"
q2a_arr_dataset$HoursBinned <- floor(q2a_arr_dataset$CRSDepTime / 100)

# "Grouping the Scheduled Departure Time into intervals or bins to analyze departure delays data set separately"
q2a_dep_dataset$Hours_Binned <- floor(q2a_dep_dataset$CRSDepTime / 100)
str(q2a_arr_dataset)
str(q2a_dep_dataset)
```


```{r}
# "Defining range and labels for binning"
range <- c(0,100,200,300,400,500,600,700,800,900,1000,1100,1200,1300,1400,1500,1600,1700,1800,1900,2000,2100,2200,2300,2400)
labels <- c('0-1','1-2','2-3','3-4','4-5','5-6','6-7','7-8','8-9','9-10','10-11','11-12','12-13','13-14','14-15','15-16','16-17','17-18','18-19','19-20','20-21','21-22','22-23','23-24')
```

```{r}
# "Binning the Scheduled Departure Time into intervals"
q2a_arr_dataset$CRSDepTime <- cut(q2a_arr_dataset$CRSDepTime, breaks = range, labels = labels)
q2a_dep_dataset$CRSDepTime <- cut(q2a_dep_dataset$CRSDepTime, breaks = range, labels = labels)

# "Changing the type of the variable CRSDepTime to category"
q2a_arr_dataset$CRSDepTime <- as.factor(q2a_arr_dataset$CRSDepTime)
q2a_dep_dataset$CRSDepTime <- as.factor(q2a_dep_dataset$CRSDepTime)
# "Rechecking data types"
str(q2a_arr_dataset)
str(q2a_dep_dataset)
```


```{r}
# "Extract unique years from the data set"
years <- unique(q2a_arr_dataset$Year)

ggplot() +
  # "Loop through each year"
  lapply(years, function(year) {
    # "Subset data for the current year"
    arr_datayearly <- q2a_arr_dataset[q2a_arr_dataset$Year == year, ]
    #" Calculate median arrival delay for each hour of the day"
    median_arrdelay_byhr <- aggregate(ArrDelay ~ CRSDepTime, data = arr_datayearly, FUN = median)
    # "Set color based on year"
    color <- ifelse(year == 2006, "red", "darkgreen")
    # "Plot median arrival delay for each hour"
    geom_line(data = median_arrdelay_byhr, aes(x = CRSDepTime, y = ArrDelay, group = 1, color = as.factor(year)), size = 1.1)
  }) +
  # "Set labels and title of the graph"
  labs(title = "Median Arrival Delay by Hour for 2006 and 2007",
       x = "Hours in each day",
       y = "Median Arrival Delay",
       color = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "right") +
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"))

```


```{r}
# "Extract unique years from the data set
years <- unique(q2a_dep_dataset$Year)

ggplot() +
  # "Loop through each year"
  lapply(years, function(year) {
    # "Subset data for the current year"
    dep_datayearly <- q2a_dep_dataset[q2a_dep_dataset$Year == year, ]
    # "Calculate median departure delay for each hour of the day"
    median_depdelay_byhr <- aggregate(DepDelay ~ CRSDepTime, data = dep_datayearly, FUN = median)
    # "Set color based on year"
    color <- ifelse(year == 2006, "red", "darkgreen")
    geom_line(data = median_depdelay_byhr, aes(x = CRSDepTime, y = DepDelay, color = as.factor(year), group = 1), size = 1)
  }) +
  labs(title = "Median Departure Delay by Hour for Different Years",
       x = "Hour",
       y = "Median Departure Delay",
       color = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "right") +
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"))

```

```{r}
# "Extract unique years from the data set"
years <- unique(q2a_arr_dataset$Year)

ggplot() +
  lapply(years, function(year) {
    yr_arrivaldata <- q2a_arr_dataset[q2a_arr_dataset$Year == year, ]
    yr_departuredata <- q2a_dep_dataset[q2a_dep_dataset$Year == year, ]
    tot_delay_byhr <- merge(
      aggregate(ArrDelay ~ CRSDepTime, data = yr_arrivaldata, FUN = median),
      aggregate(DepDelay ~ CRSDepTime, data = yr_departuredata, FUN = median),
      by = "CRSDepTime"
    ) %>%
      transmute(CRSDepTime, TotalDelay = ArrDelay + DepDelay)
    color <- ifelse(year == 2006, "red", "darkgreen")
    geom_line(data = tot_delay_byhr, aes(x = CRSDepTime, y = TotalDelay, color = as.factor(year), group = 1), size = 0.9)
  }) +
  labs(title = "Total Delay by Hour for Different Years",
       x = "Hour",
       y = "Total Delay",
       color = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "right") +
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"))
```

*************************************************************
*"(iii) Analysis for the Best day of the week to travel"*
*************************************************************


```{r}

color_2006 <- "#8474c7"
color_2007 <- "#71bbee"

years <- unique(q2a_arr_dataset$Year)

plot_data <- data.frame()

for (year in years) {
  yr_arrdata <- q2a_arr_dataset[q2a_arr_dataset$Year == year, ]
  med_arr_delay_byhr <- aggregate(ArrDelay ~ DayOfWeek, data = yr_arrdata, FUN = median)
  med_arr_delay_byhr$Year <- year
  plot_data <- rbind(plot_data, med_arr_delay_byhr)
}

ggplot(plot_data, aes(x = as.factor(DayOfWeek), y = ArrDelay, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.4), width = 0.4) +
  scale_fill_manual(values = c(color_2006, color_2007)) +
  labs(x = "Day of Week", y = "Delay Time (minutes)", title = "Median Arrival Delays by Week", fill = "Year") +
  scale_x_discrete(labels = c("MON", "TUE", "WED", "THU", "FRI", "SAT", "SUN")) +
  theme_minimal() +
  theme(legend.position = "top")
```


```{r}
color_2006 <- "#ae8fdc"
color_2007 <- "#86ddf3"

years <- unique(q2a_dep_dataset$Year)

plot_data <- data.frame()

for (year in years) {
  yr_depdata <- q2a_dep_dataset[q2a_dep_dataset$Year == year, ]
  med_depdelay_byhr <- aggregate(DepDelay ~ DayOfWeek, data = yr_depdata, FUN = median)
  med_depdelay_byhr$Year <- year
  plot_data <- rbind(plot_data, med_depdelay_byhr)
}

# "the plot"
ggplot(plot_data, aes(x = as.factor(DayOfWeek), y = DepDelay, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.4), width = 0.4) +
  scale_fill_manual(values = c(color_2006, color_2007)) +
  labs(x = "Day of Week", y = "Delay Time (minutes)", title = "Median Departure Delays by Week", fill = "Year") +
  scale_x_discrete(labels = c("MON", "TUE", "WED", "THU", "FRI", "SAT", "SUN")) +
  theme_minimal() +
  theme(legend.position = "top")
```

```{r}
color_2006 <- "#d395e0"
color_2007 <- "#feb2da"

years <- unique(q2a_arr_dataset$Year)

plot_data <- data.frame()

for (year in years) {
  yr_arrdata <- q2a_arr_dataset[q2a_arr_dataset$Year == year, ]
  yr_depdata <- q2a_dep_dataset[q2a_dep_dataset$Year == year, ]
  tot_delay_byhr <- aggregate(ArrDelay ~ DayOfWeek, data = yr_arrdata, FUN = median) +
                    aggregate(DepDelay ~ DayOfWeek, data = yr_depdata, FUN = median)
  tot_delay_byhr$Year <- year
  plot_data <- rbind(plot_data, tot_delay_byhr)
}

ggplot(plot_data, aes(x = as.factor(DayOfWeek), y = ArrDelay, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.4), width = 0.4) +
  scale_fill_manual(values = c(color_2006, color_2007)) +
  labs(x = "Day of Week", y = "Delay Time (minutes)", title = "Median Total Delays by Week", fill = "Year") +
  scale_x_discrete(labels = c("MON", "TUE", "WED", "THU", "FRI", "SAT", "SUN")) +
  theme_minimal() +
  theme(legend.position = "top")

```

```{r}
# "Defining the days in a week"
days_inweek <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday')

# "Loop through each year"
for (year in unique(q2a_arr_dataset$Year)) {
  plt <- ggplot()
  
  
  for (i in 1:length(days_inweek)) {
    # "Filter data for the current year and weekday"
    yearly_arr <- q2a_arr_dataset[q2a_arr_dataset$Year == year & q2a_arr_dataset$DayOfWeek == i, ]
    yearly_dep <- q2a_dep_dataset[q2a_dep_dataset$Year == year & q2a_dep_dataset$DayOfWeek == i, ]
    
    # "Calculate median arrival delay for each hour"
    median_arrdelay <- aggregate(ArrDelay ~ CRSDepTime, data = yearly_arr, FUN = median)
    
    # "Calculate median departure delay for each hour"
    median_depdelay <- aggregate(DepDelay ~ CRSDepTime, data = yearly_dep, FUN = median)
    
    # "Calculate total median delay for each hour"
    tot_median_delay <- median_arrdelay$ArrDelay + median_depdelay$DepDelay
    
    # "Combine data for plotting"
    plot_data <- data.frame(Hour = median_arrdelay$CRSDepTime, Total_Median_Delay = tot_median_delay, Weekday = days_inweek[i])
    
    # "Add lines to the plot for each weekday"
    plt <- plt + geom_line(data = plot_data, aes(x = Hour, y = Total_Median_Delay, group = Weekday, color = Weekday), 
                           size = ifelse(year == 2006, 0.8, 1))
  }
  
  # "Set plot "
  plt <- plt + labs(title = paste("Total Median Delay by Hour for Each Day of the Week -", year),
                    x = "Hour of Day",
                    y = "Total Median Delay (minutes)") +
    scale_color_brewer(palette = "Set1") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(legend.position = "bottom") +
    theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"))

  print(plt)
}

```


```{r}
# "Define days of the week"
days_inweek <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday')

# "Create a new plot"
plt <- ggplot()

# "Define colors for each combination of weekday and year"
line_colors <- c("Thursday_2006" = "red", "Friday_2006" = "blue", "Wednesday_2007" = "green",
                 "Thursday_2007" = "orange", "Sunday_2007" = "purple")

# "Loop through each year"
for (year in c(2006, 2007)) {
  # "Loop through selected days of the week"
  for (i in 1:length(days_inweek)) {
    if ((year == 2006 && i %in% c(4, 5)) || (year == 2007 && i %in% c(3, 4, 7))) {
      # "Filter data for the current year and selected days of the week"
      year_arrdata <- q2a_arr_dataset[q2a_arr_dataset$Year == year & q2a_arr_dataset$DayOfWeek == i, ]
      # "Calculate median arrival delay for each hour"
      median_arrdelay <- aggregate(ArrDelay ~ CRSDepTime, data = year_arrdata, FUN = median)
      # "Combine data for plotting"
      plot_data <- data.frame(Hour = median_arrdelay$CRSDepTime, Median_Arrival_Delay = median_arrdelay$ArrDelay, 
                              Year = factor(year), Weekday = days_inweek[i], Category = paste(days_inweek[i], year, sep = "_"))
      plt <- plt + geom_line(data = plot_data, aes(x = Hour, y = Median_Arrival_Delay, 
                                                   color = Category, group = Category), 
                             size = 1)
    }
  }
}

# "Set plot" 
plt <- plt + labs(title = "Median Arrival Delay by Hour for Selected Days of the Week",
                  x = "Hour of Day",
                  y = "Median Arrival Delay (minutes)",
                  color = "Category") +
  scale_color_manual(values = line_colors) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 9),
        legend.key.size = unit(2, "lines"),
        legend.box.margin = margin(0, 10, 0, 10)) +
  guides(color = guide_legend(override.aes = list(size = 0.5)))

print(plt)

```


```{r}

str(q2a_dep_dataset)
summary(is.na(q2a_dep_dataset))

```

```{r}
# "Convert CRSDepTime to numeric"
q2a_dep_dataset$CRSDepTime <- as.numeric(levels(q2a_dep_dataset$CRSDepTime))[q2a_dep_dataset$CRSDepTime]

# "Check the structure of the data set again"
str(q2a_dep_dataset)

```

```{r}
# "Check unique values of CRSDepTime before conversion"
unique_values <- unique(q2a_dep_dataset$CRSDepTime)
print(unique_values)
```

```{r}
# "Remove rows with NA values in CRSDepTime"
q2a_dep_dataset <- q2a_dep_dataset[!is.na(q2a_dep_dataset$CRSDepTime), ]

# "check the structure of the data set again"
str(q2a_dep_dataset)

```


```{r}

# "Define days of the week"
days_inweek <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday')

# "Define custom colors"
colors <- c('#fa3b5a', '#028886')

# "Create a new plot"
plt <- ggplot()

# "Loop through each year"
for (idx in 1:2) {
  year <- c(2006, 2007)[idx]
  
  # "Loop through selected days of the week"
  for (i in 1:length(days_inweek)) {
    # "Filter data for the current year and selected weekdays"
    yr_depdata <- q2a_dep_dataset[q2a_dep_dataset$Year == year & q2a_dep_dataset$DayOfWeek == i, ]
    
    # "Check if there are any rows left after filtering"
    if (nrow(yr_depdata) > 0) {
      # "Remove missing values"
      yr_depdata <- na.omit(yr_depdata)
      
      # "Convert CRSDepTime to numeric if it's a factor"
      if (is.factor(yr_depdata$CRSDepTime)) {
        yr_depdata$CRSDepTime <- as.numeric(as.character(yr_depdata$CRSDepTime))
      }
      
      # "Check if there are any rows left after removing missing values"
      if (nrow(yr_depdata) > 0) {
        # "Calculate median departure delay for each hour"
        med_depdelay <- tapply(yr_depdata$DepDelay, yr_depdata$CRSDepTime, median, na.rm = TRUE)
        
        # "Convert the result to a data frame"
        med_depdelay_df <- data.frame(CRSDepTime = as.numeric(names(med_depdelay)), DepDelay = med_depdelay)
        
        
        plt <- plt + geom_line(data = med_depdelay_df, aes(x = CRSDepTime, y = DepDelay, group = 1), 
                               color = colors[idx], size = 2) +
                labs(title = paste("Median Departure Delay by Hour for Selected Days of the Week"), 
                     x = "Hour of Day", y = "Median Departure Delay (minutes)")
      }
    }
  }
}

# " the plot"
plt <- plt +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        legend.key.size = unit(2, "lines"),
        legend.box.margin = margin(0, 10, 0, 10)) +
  scale_x_continuous(breaks = seq(0, 2400, by = 100))

print(plt)
```


END






























."
